---
title: "Climate Figures"
output:
  pdf_document: default
  html_document: default
date: "2026-04-01"
---

# Libraries Data Cleaning and Setup 

```{r message=FALSE}

library(tidyverse)
library(dplyr)
library(sf)
library(lubridate)
library(mgcv) 
library(lme4)
library(reshape2)
library(readr)


```

```{r}

climate <- read_csv("beetles_with_climate.csv")
head(climate)

```

```{r}

climate_clean <- climate %>%
  mutate(
    SURVEY_YEAR = as.numeric(SURVEY_YEA),
    ACRES = as.numeric(ACRES),
    PERCENT_MID = as.numeric(PERCENT_MI)
  ) %>% 
  select(-CREATED_DA, -MODIFIED_D,
         -FEATURE_US, -OBSERVATIO, 
         -NOTES, -COLLECTION, 
         -SNAPGRID_R, -SNAPGRID_C, 
         -HOST_GROUP, -HOST_CODE, 
         -LABEL, -DAMAGE_ARE, 
         -HOST_GRO_1, -DCA_CODE, 
         -OBSERVAT_1, -AREA_TYPE, 
         -STATUS, -GRP, -OBSERVAT_2, 
         -DATA_SOURC, -IDS_DATA_S, 
         -SHAPE_Leng, -SHAPE_Area, 
         -LEGACY_PAT, -LEGACY_P_1, 
         -LEGACY_TPA, -LEGACY_NO_, 
         -LEGACY_FOR, -LEGACY_F_1) %>% 
  filter(DCA_COMMON == "mountain pine beetle", HOST == "ponderosa pine", DAMAGE_T_1 == "Mortality")
head(climate_clean)
```

# Non-Climate Plots

```{r}

total_affected <- climate_clean %>%
  group_by(SURVEY_YEA) %>%
  summarise(total_acres = sum(ACRES, na.rm = TRUE)) %>%
  ggplot(aes(x = SURVEY_YEA, y = total_acres)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point() +
  labs(title = "Total Affected Area Over Time",
       x = "Year", y = "Total Acres") +
  theme_minimal()

total_affected
```

**Summary**

\- shows how the total area of forest affected by mortality changes over time

\- each point represents the summed acres impacted in a given survey year

\- line highlights trends in outbreak dynamics

\- huge spike and then it flat lines (which may be due to lack of data)

```{r}

mortality <- climate_clean %>%
  group_by(SURVEY_YEA) %>%
  summarise(mean_mortality = mean(PERCENT_MI, na.rm = TRUE)) %>%
  ggplot(aes(SURVEY_YEA, mean_mortality)) +
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "Mean Mortality Severity Over Time",
       y = "Percent Mortality", x = "Survey Year") +
  theme_minimal()
mortality

mortality_zoom <- climate_clean %>%
  filter(SURVEY_YEA >= 2010 & SURVEY_YEA <= 2025) %>%
  group_by(SURVEY_YEAR) %>%
  summarise(mean_mortality = mean(PERCENT_MI, na.rm = TRUE)) %>%
  ggplot(aes(x = SURVEY_YEAR, y = mean_mortality)) +
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "Mean Mortality Severity (2010-2025)",
       y = "Percent Mortality", x = "Survey Year") +
  theme_minimal()
mortality_zoom
```

**Summary**

\- average mortality severity of ponderosa pines from 2010 to 2025

\- highlights year-to-year trends, whether the outbreak was intensifying, stabilizing, or declining during that period

\- shape looks decent

\- the second one is the same just zoomed in

```{r}

ggplot(climate_clean, aes(x = ACRES)) +
  geom_histogram(bins = 40, fill = "forestgreen") +
  scale_x_log10() +
  labs(title = "Distribution of Affected Area (log scale)", x = "Acres") +
  theme_minimal()
```

**Summary**

\- distribution of individual affected areas across all observations

\- log scale to account for the wide range of sizes

\- reveals that most mortality events impact relatively small areas

```{r}

ggplot(climate_clean, aes(x = PERCENT_MI)) +
  geom_histogram(bins = 30, fill = "orange") +
  labs(title = "Distribution of Mortality Severity", x = "Percent Mortality") +
  theme_minimal()

# Percent Mid = Affected range mid point
```

**Summary**

\- shows how mortality severity (PERCENT_MID) is distributed

\- indicates that most observations have low to moderate mortality

\- high-severity events are less common

```{r}

ggplot(climate_clean %>% filter(!is.na(PERCENT_MI)), 
       aes(x = factor(SURVEY_YEA), y = PERCENT_MI)) +
  geom_boxplot() +
  labs(title = "Mortality by Year",
       x = "Year", y = "Percent Mortality") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

**summary**

\- something feels wrong here but we still have the right trend

\- distribution of mortality severity each year

```{r}
ggplot(climate_clean, aes(x = ACRES, y = PERCENT_MI)) +
  scale_x_log10() +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Area vs Mortality Severity", x = "Acres", y = "Percent Mortality") +
  theme_minimal()
```

# Climate Based Plots

```{r}
ggplot(climate_clean, aes(x = tmean, y = log(ACRES))) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Mean Annual Temperature vs. Pine Beetle Mortality",
    x = "Mean Annual Temperature (°C)",
    y = "Log of Mortality Area (Acres)"
  ) + 
  theme_minimal()

ggplot(climate_clean, aes(x = tmin, y = log(ACRES))) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Minimum Temperature (Winter Conditions) vs. Pine Beetle Mortality",
    x = "Minimum Temperature (°C)",
    y = "Log of Mortality Area (Acres)"
  ) + 
  theme_minimal()

ggplot(climate_clean, aes(x = tmax, y = log(ACRES))) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Maximum Temperature vs. Pine Beetle Mortality",
    x = "Maximum Temperature (°C)",
    y = "Log of Mortality Area (Acres)"
  ) + 
  theme_minimal()
```

**Summary** \
- mortality higher when temperature lower?? am I reading that right

```{r}
ggplot(climate_clean, aes(x = tmean, y = PERCENT_MI)) +
  geom_smooth() + 
  labs(
    title = "Mean Annual Temperature vs. Percent of Area Affected",
    x = "Mean Annual Temperature (°C)",
    y = "Percent of Area Affected"
  ) + 
  theme_minimal()

ggplot(climate_clean, aes(x = tmin, y = PERCENT_MI)) +
  geom_smooth() + 
  labs(
    title = "Minimum Temperature vs. Percent of Area Affected",
    x = "Minimum Temperature (°C)",
    y = "Percent of Area Affected"
  ) + 
  theme_minimal()

ggplot(climate_clean, aes(x = tmax, y = PERCENT_MI)) +
  geom_smooth() + 
  labs(
    title = "Maximum Temperature vs. Percent of Area Affected",
    x = "Maximum Temperature (°C)",
    y = "Percent of Area Affected"
  ) + 
  theme_minimal()
```

**Summary**

-   I think these show what the first three were trying to show better

-   starts low peaks and then flattens out

-   I think the minimum temperature one is the most interesting

```{r}

ggplot(climate_clean, aes(x = ppt, y = PERCENT_MI)) +
  geom_smooth(color = "blue") +
  labs(
    title = "Annual Precipitation vs. Percent of Area Affected by Pine Beetle Mortality",
    x = "Annual Precipitation (mm)",
    y = "Percent of Area Affected"
  ) + 
  theme_minimal()
```

**Summary**

-   notable peak around 14mm

-   dry conditions favor beetles

-   This one is good

```{r}

ggplot(climate_clean, aes(x = vpdmax, y = PERCENT_MI)) +
  geom_smooth(color = "blue") +
  labs(
    title = "Maximum Vapor Pressure Deficit vs. Percent of Area Affected",
    x = "Maximum Vapor Pressure Deficit (hPa)",
    y = "Percent of Area Affected "
  )
```

**Summary:**

-   peak \~14mm

-   lots of room for error around the edges

-   suggests that moderate atmospheric dryness may favor beetle activity?

```{r}

library(viridis)

ggplot(climate_clean, aes(x = tmax, y = ppt)) +
  geom_bin2d() +
  scale_fill_viridis_c(option = "C", name = "Observation Count") +
  labs(
    title = "Relationship Between Maximum Temperature and Annual Precipitation",
    x = "Maximum Temperature (°C)",
    y = "Annual Precipitation (mm)"
  ) +
  theme_minimal(base_size = 14)
```

**Summary :**

-   There seems to be a temperature and percip sweet spot for beetle activity

# Models

```{r}
climate_model <- climate_clean %>%
  filter(!is.na(ACRES), !is.na(tmean), !is.na(ppt), !is.na(vpdmax)) %>%
  mutate(
    log_acres = log(ACRES),
    SURVEY_YEAR = as.numeric(SURVEY_YEAR)
  )
```

```{r}
lm <- lm(
  log(ACRES) ~ 
    tmean + 
    tmin + 
    tmax + 
    ppt + 
    vpdmax,
  data = climate_clean
)

summary(lm)
plot(lm, pages = 4)

climate_clean$pred_log_acres <- predict(lm, newdata = climate_clean)

ggplot(climate_clean, aes(x = log(ACRES), y = pred_log_acres)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Observed vs. Predicted Pine Beetle Mortality Area",
    x = "Observed Mortality Area (log acres)",
    y = "Predicted Mortality Area (log acres)"
  ) +
  theme_minimal()
```

```{r}
gam_model <- gam(
  log_acres ~ 
    s(tmean) +
    s(ppt) +
    s(vpdmax) +
    s(tmin) +
    s(SURVEY_YEAR),
  data = climate_model,
  method = "REML"
)

summary(gam_model)
plot(gam_model, pages = 5)
```

```{r}
library(gratia)

gam_model2 <- gam(
  log_acres ~ 
    te(tmean, ppt) +  
    s(vpdmax) +
    s(SURVEY_YEAR),
  data = climate_model,
  method = "REML"
)

summary(gam_model2)

draw(gam_model2, select = 1) + 
  labs(
    title = "Predicted Pine Beetle Mortality",
    x = "Mean Annual Temperature (°C)",
    y = "Annual Precipitation (mm)",
    fill = "Predicted Mortality\n(log acres)"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
gamm(
  log_acres ~ te(tmean, ppt) + s(vpdmax) + s(SURVEY_YEAR),
  random = list(REGION_ID = ~1),
  data = climate_model
)

gamm_model <- gamm(
  log_acres ~ 
    s(tmean) +
    s(ppt) +
    s(vpdmax) +
    s(tmin),
  random = list(SURVEY_YEAR = ~1),
  data = climate_model,
  method = "REML"
)

par(mfrow = c(2, 2))  # bigger panels
plot(gamm_model$gam,
     cex = 1.2,
     lwd = 2, pages = 4)
```
