---
title: "SSRD FEM Data"
output:
  pdf_document: default
  html_document: default
date: "2026-03-24"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(sf)
library(lubridate)
library(mgcv) 
library(lme4)  
library(reshape2)
```

```{r message=FALSE}
library(readr)
FEM_Data <- read_csv("FEM Data.csv")
```

```{r}

PondoMPB_Data <- FEM_Data %>% 
  filter(HOST == "ponderosa pine", DCA_COMMON_NAME == "mountain pine beetle", DAMAGE_TYPE == "Mortality") %>% 
  mutate(
    SURVEY_YEAR = as.numeric(SURVEY_YEAR),
    ACRES = as.numeric(ACRES),
    PERCENT_MID = as.numeric(PERCENT_MID)
  ) %>% 
  select(-CREATED_DATE, -MODIFIED_DATE,
    -FEATURE_USER_ID, -OBSERVATION_USER_ID,
    -NOTES, -COLLECTION_MODE,
    -SNAPGRID_ROW, -SNAPGRID_COLUMN, -HOST_GROUP, -HOST_GROUP_CODE, -PERCENT_AFFECTED_CODE, -LABEL, -matches("^X_clim_.*(min|max|count|sum)$"))

```

```{r}
total_affected <- PondoMPB_Data %>%
  group_by(SURVEY_YEAR) %>%
  summarise(total_acres = sum(ACRES, na.rm = TRUE)) %>%
  ggplot(aes(x = SURVEY_YEAR, y = total_acres)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point() +
  labs(title = "Total Affected Area Over Time",
       x = "Year", y = "Total Acres") +
  theme_minimal()

total_affected
```

**Summary**

\- shows how the total area of forest affected by mortality changes over
time

\- each point represents the summed acres impacted in a given survey
year

\- line highlights trends in outbreak dynamics

\- huge spike and then it flat lines (which may be due to lack of data)

```{r}
mortality <- PondoMPB_Data %>%
  group_by(SURVEY_YEAR) %>%
  summarise(mean_mortality = mean(PERCENT_MID, na.rm = TRUE)) %>%
  ggplot(aes(SURVEY_YEAR, mean_mortality)) +
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "Mean Mortality Severity Over Time",
       y = "Percent Mortality") +
  theme_minimal()
mortality

mortality <- PondoMPB_Data %>%
  filter(SURVEY_YEAR >= 2010 & SURVEY_YEAR <= 2025) %>%
  group_by(SURVEY_YEAR) %>%
  summarise(mean_mortality = mean(PERCENT_MID, na.rm = TRUE)) %>%
  ggplot(aes(x = SURVEY_YEAR, y = mean_mortality)) +
  geom_line(color = "blue") +
  geom_point() +
  labs(title = "Mean Mortality Severity (2010-2025)",
       y = "Percent Mortality") +
  theme_minimal()

mortality
```

**Summary**

\- average mortality severity of ponderosa pines from 2010 to 2025

\- highlights year-to-year trends, whether the outbreak was
intensifying, stabilizing, or declining during that period

\- shape looks decent

```{r}
ggplot(PondoMPB_Data, aes(x = ACRES)) +
  geom_histogram(bins = 40, fill = "forestgreen") +
  scale_x_log10() +
  labs(title = "Distribution of Affected Area (log scale)") +
  theme_minimal()
```

**Summary** - distribution of individual affected areas across all
observations - log scale to account for the wide range of sizes -
reveals that most mortality events impact relatively small areas

```{r}
ggplot(PondoMPB_Data, aes(x = PERCENT_MID)) +
  geom_histogram(bins = 30, fill = "orange") +
  labs(title = "Distribution of Mortality Severity") +
  theme_minimal()

# Percent Mid = Affected range mid point
```

**Summary**

\- shows how mortality severity (PERCENT_MID) is distributed

\- indicates that most observations have low to moderate mortality

\- high-severity events are less common

```{r}
ggplot(PondoMPB_Data %>% filter(!is.na(PERCENT_MID)), 
       aes(x = factor(SURVEY_YEAR), y = PERCENT_MID)) +
  geom_boxplot() +
  labs(title = "Mortality by Year",
       x = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```

**summary**

\- something feels wrong here but we still have the right trend

\- distribution of mortality severity each year

```{r}
ggplot(PondoMPB_Data, aes(x = ACRES, y = PERCENT_MID)) +
  scale_x_log10() +
  geom_smooth(method = "loess", color = "red") +
  labs(title = "Area vs Mortality Severity") +
  theme_minimal()
```

**Summary**

\- relationship between the size of affected areas (log-scaled) and
mortality severity

\- red curve = in general, larger affected areas tend to be associated
with slightly higher mortality

\- relationship is nonlinear and variable

```{r}
climate_vars <- names(PondoMPB_Data) %>%
  stringr::str_subset("^X_clim_.*mean$")

correlations <- sapply(PondoMPB_Data[climate_vars], function(x) {
  cor(x, PondoMPB_Data$PERCENT_MID, use = "complete.obs")
})

sort(correlations, decreasing = TRUE)

top_vars <- names(sort(abs(correlations), decreasing = TRUE))[1:6]

df_top <- PondoMPB_Data %>%
  pivot_longer(
    cols = all_of(top_vars),
    names_to = "climate_var",
    values_to = "climate_value"
  )

ggplot(df_top, aes(x = climate_value, y = PERCENT_MID)) +
  geom_smooth(method = "gam", formula = y ~ s(x), color = "blue") +
  facet_wrap(~ climate_var, scales = "free_x", ncol = 2) +
  theme_minimal() +
  labs(
    title = "Top Climate Drivers of Mortality",
    x = "Climate Variable",
    y = "Percent Mortality"
  )

```

**Summary**

\- six climate variables most strongly correlated with mortality

\- I hate these climate variables and have no idea what they mean

\- each panel represents one of the 6

\- GAM smoothed

\- mortality responds differently across the range of each climate
driver

```{r}
climate_vars <- PondoMPB_Data %>%
  select(starts_with("X_clim")) %>%
  select(where(is.numeric))

cor_mat <- cor(climate_vars, use = "complete.obs")


cor_df <- melt(cor_mat)

ggplot(cor_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Climate Variable Correlation Matrix")

```

**Summary**

\- pairwise correlations among all mean numeric climate variables

\- colors = strength and direction

\- I know the axis suck I can make them better if I have better names
for the climate variables

```{r}
severity_overtime <- PondoMPB_Data %>%
  count(SURVEY_YEAR, PERCENT_AFFECTED) %>%
  ggplot(aes(x = SURVEY_YEAR, y = n, fill = PERCENT_AFFECTED)) +
  geom_area() +
  labs(title = "Severity Classes Over Time",
       y = "Count") +
  theme_minimal()
severity_overtime

severity_overtime_zoom <- PondoMPB_Data %>%
  filter(!is.na(PERCENT_AFFECTED),
         PERCENT_AFFECTED != "") %>%
  count(SURVEY_YEAR, PERCENT_AFFECTED) %>%
  ggplot(aes(x = SURVEY_YEAR, y = n, fill = PERCENT_AFFECTED)) +
  geom_area() +
  labs(title = "Severity Classes Over Time",
       y = "Count") +
  theme_minimal()
severity_overtime_zoom



```

**Summary:**

\- how number of observations in each mortality severity class changes
over time

-   I zoomed in on just the part we have the right data for to make it
    readable and look prettier

\- I think with the rest of our data these trends make sense

# Modeling

```{r}
lm_data <- PondoMPB_Data %>%
  select(PERCENT_MID, all_of(top_vars)) %>%
  filter(!is.na(PERCENT_MID)) %>%
  drop_na()

lm_model <- lm(PERCENT_MID ~ ., data = lm_data)
summary(lm_model)
```

```{r}

lm_data$predicted <- predict(lm_model, newdata = lm_data)

ggplot(lm_data, aes(x = predicted, y = PERCENT_MID)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Linear Model: Observed vs Predicted Mortality",
       x = "Predicted Percent Mortality",
       y = "Observed Percent Mortality") +
  theme_minimal()

ggplot(lm_data, aes(x = predicted, y = PERCENT_MID)) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(title = "Linear Model: Observed vs Predicted Mortality",
       x = "Predicted Percent Mortality",
       y = "Observed Percent Mortality") +
  theme_minimal()

lm_long <- lm_data %>%
  pivot_longer(cols = all_of(top_vars), names_to = "predictor", values_to = "value")

ggplot(lm_long, aes(x = value, y = PERCENT_MID)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(title = "Partial Effects of Climate Predictors on Mortality",
       x = "Predictor Value", y = "Percent Mortality") +
  theme_minimal()
```

```{r}
gam_model <- gam(PERCENT_MID ~ s(X_clim_dat23mean) + s(X_clim_dat97mean) +
                   s(X_clim_dat16mean) + s(X_clim_dat20mean) +
                   s(X_clim_dat99mean) + s(X_clim_dat14mean),
                 data = PondoMPB_Data,
                 method = "REML",
                 na.action = na.omit)

summary(gam_model)

par(mfrow = c(2, 3), mar = c(5, 5, 3, 2))
plot(gam_model, rug = TRUE, shade = TRUE, pages = 1)
```
