---
title: "gooningToBeetles"
output: pdf_document
date: "2026-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#I LOVE TIDYVERSE
```{r}
library(tidyverse)
```

#dumbass beetles
```{r}
beets = read_csv('~/DATA399/SemesterGroupProject/rooseveltBeetles.csv')
```

#Just need ponderosas
```{r}
beets = beets %>%
  filter(HOST == 'ponderosa pine')
```
#What's going up there in that noggin of yours??
```{r}
colSums(is.na(beets))
nrow(beets)
```

#no dawg these rows shit (too many nas can't be bothered to deal with)
```{r}
beets =  subset(beets, select = -c(HOST_GROUP_CODE, PERCENT_AFFECTED_CODE, PERCENT_MIN,
                   PERCENT_MAX, PERCENT_MID, SNAPGRID_ROW, SNAPGRID_COLUMN))
```

#Mmkay how we lookin now??
```{r}
is(beets$)
```



#uhh let's see here trying to steal the internet code to import climate data mmmmm climate data

#No don't runn it againnn (jk just run it once but it kinda slow)
```{}
install.packages("prism")
library(prism)

#Thingy going in here
prism_set_dl_dir("~/DATA399/SemesterGroupProject/prism_data")

# Download annual precip and temp for your year range (1997-2024)
get_prism_annual(type = "ppt", years = 1997:2024, keepZip = FALSE)
get_prism_annual(type = "tmean", years = 1997:2024, keepZip = FALSE)
```

#I really don't know what this does..
```{}
#install.packages('sf')
#install.packages('terra')
library(prism)
library(terra)

prism_set_dl_dir("~/DATA399/SemesterGroupProject/prism_data")

get_prism_annual(type = "ppt", years = 1997:2024, keepZip = FALSE)

# Replace pd_stack with this:
prism_files <- prism_archive_subset("ppt", "annual", years = 1997:2024,
                                    resolution = '4k')
prism_paths <- pd_to_file(prism_files)  # gets the actual .bil file paths

# Stack with terra directly
prism_stack <- rast(prism_paths)

# Then extract as before
climate_vals <- terra::extract(prism_stack, beets, fun = mean)
```

```{}
library(sf)
library(dplyr)

gpkg_files <- list.files("~/DATA399/SemesterGroupProject/idk some output", pattern = "\\.gpkg$", full.names = TRUE)

# Read all files
layers <- lapply(gpkg_files, st_read)

# Drop geometry from all but the first, then join by a shared ID
combined <- layers[[1]]

for (i in 2:length(layers)) {
  # drop geometry before joining
  new_cols <- st_drop_geometry(layers[[i]])
  # only keep the ID column + the new climate columns
  new_cols <- select(new_cols, OBJECTID, contains("mean"))
  combined <- left_join(combined, new_cols, by = "OBJECTID")
}

st_write(combined, "beetles_with_climate_wide.gpkg")
```


