---
title: "ToBeetles"
output: pdf_document
date: "2026-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#I LOVE TIDYVERSE
```{r}
library(tidyverse)
```

#dumbass beetles
```{r}
beets = read_csv('~/DATA399/SemesterGroupProject/rooseveltBeetles.csv')
```

#Just need ponderosas
```{r}
beets = beets %>%
  filter(HOST == 'ponderosa pine')
```
#What's going up there in that noggin of yours??
```{r}
colSums(is.na(beets))
nrow(beets)

sort(unique(beets$SURVEY_YEAR))
```

#no dawg these rows shit (too many nas can't be bothered to deal with)
```{r}
beets =  subset(beets, select = -c(HOST_GROUP_CODE, PERCENT_AFFECTED_CODE, PERCENT_MIN,
                   PERCENT_MAX, PERCENT_MID, SNAPGRID_ROW, SNAPGRID_COLUMN))
```

#Mmkay how we lookin now??
```{r}
is(beets$)
```


#downloads all the prism data, and I mean allllll of the prism data
```{r}
#install.packages("prism")
library(prism)

#Thingy going in here
prism_set_dl_dir("~/DATA399/SemesterGroupProject/prism_data2")

#We're downloading total precipitation
get_prism_annual(type = "ppt", years = 1997:2024, keepZip = FALSE)
#mean temperature
get_prism_annual(type = "tmean", years = 1997:2024, keepZip = FALSE)
#max temperature
get_prism_annual(type = "tmax", years = 1997:2024, keepZip = FALSE)
#min temperature
get_prism_annual(type = "tmin", years = 1997:2024, keepZip = FALSE)
#min vapor pressure deficit
get_prism_annual(type = "vpdmin", years = 1997:2024, keepZip = FALSE)
#max vapor pressure deficit
get_prism_annual(type = "vpdmax", years = 1997:2024, keepZip = FALSE)
#mean dew point temperature
get_prism_annual(type = "tdmean", years = 1997:2024, keepZip = FALSE)
```

#uhh let's see here trying to steal the internet code to import climate data mmmmm climate data

#I really don't know what this does..


```{r}
library(sf)
library(terra)
library(dplyr)

# Load and fix CRS
beetles <- st_read("~/DATA399/SemesterGroupProject/damage_areas_new_crs.shp")
# CRS is correctly EPSG:5070 now, just reproject to PRISM's CRS
beetles_proj <- st_transform(beetles, crs = "EPSG:4269")

# KEY FIX: extract using centroids, not polygon boundaries
# 99.6% of polygons are smaller than one PRISM 4km cell,
# so polygon extraction almost always returns NA
beetles_centroids <- st_centroid(beetles_proj)

variables <- c("tmean", "tmax", "tmin", "tdmean", "ppt", "vpdmin", "vpdmax")
prism_dir <- "C:/Users/ochin/OneDrive/Documents/DATA399/SemesterGroupProject/prism_data2"

get_bil_path <- function(var, year, dir) {
  pattern <- paste0("prism_", var, "_us_25m_", year)
  folder  <- list.files(dir, pattern = pattern, full.names = TRUE)
  if (length(folder) == 0) return(character(0))
  bil <- list.files(folder, pattern = "\\.bil$", full.names = TRUE)
  return(bil)
}

results <- st_drop_geometry(beetles_proj)
for (var in variables) {
  results[[var]] <- NA_real_
}

for (year in sort(unique(beetles_proj$SURVEY_YEA))) {
  row_idx          <- which(beetles_proj$SURVEY_YEA == year)
  centroids_year   <- beetles_centroids[row_idx, ]

  for (var in variables) {
    bil_path <- get_bil_path(var, year, prism_dir)

    if (length(bil_path) == 0) {
      message(paste("Missing raster:", var, year))
      next
    }

    r         <- rast(bil_path)
    extracted <- terra::extract(r, vect(centroids_year), method = "simple", bind = FALSE)
    results[[var]][row_idx] <- extracted[[2]]
  }

  message(paste("Done: year", year))
}

beetles_climate <- st_sf(results, geometry = st_geometry(beetles_proj))

if ("fid" %in% names(beetles_climate)) {
  beetles_climate <- beetles_climate %>% rename(orig_fid = fid)
}

st_write(
  beetles_climate,
  "C:/Users/ochin/Downloads/School/beetles_with_climate.gpkg",
  delete_dsn = TRUE
)
```






